%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                          Module Description                           %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \index{Parameters \textit{(module)}|(}
\section{Module Parameters}

    \label{Parameters}
This is the configuration file for the multistate simulation code. All
available configuration parameters can be set here. This file is imported
from the simulation code. For configuring the code, just modifiy the values
here.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               Variables                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

  \subsection{Variables}

    \vspace{-1cm}
\hspace{\varindent}\begin{longtable}{|p{\varnamewidth}|p{\vardescrwidth}|l}
\cline{1-2}
\cline{1-2} \centering \textbf{Name} & \centering \textbf{Description}& \\
\cline{1-2}
\endhead\cline{1-2}\multicolumn{3}{r}{\small\textit{continued on next page}}\\\endfoot\cline{1-2}
\endlastfoot\raggedright a\-l\-g\-o\-r\-i\-t\-h\-m\- & \raggedright The algorithm used for time propagation, can be one of: \texttt{fourier}
          {\textbar} \texttt{hagedorn} {\textbar} \texttt{multihagedorn}.

&\\
\cline{1-2}
\raggedright p\-o\-t\-e\-n\-t\-i\-a\-l\- & \raggedright The potential $V\ofs{x}$ used in the simulation. See the
          \textit{PotentialLibrary} for available potentials.

&\\
\cline{1-2}
\raggedright T\- & \raggedright Perform a simulation in the time interval $[0, T]$.

&\\
\cline{1-2}
\raggedright d\-t\- & \raggedright Duration of a single time step $\tau$.

&\\
\cline{1-2}
\raggedright e\-p\-s\- & \raggedright The parameter $\varepsilon$ in the semiclassical
          scaling.

&\\
\cline{1-2}
\raggedright d\-e\-l\-t\-a\- & \raggedright A variable that is used in the definition of some potentials.

&\\
\cline{1-2}

\raggedright c\-o\-e\-f\-f\-i\-c\-i\-e\-n\-t\-s\- & \raggedright A list of $N$ lists of $(k,c_k)$ tuples that set the
          coefficient $c_k$ of the basis function
          $\phi_k$. The $i$-th list contains the
          coefficients $c^i$ of the component
          $\Phi_i$ of the initial wavepacket
          $\Psi$.

&\\
\cline{1-2}
\raggedright p\-a\-r\-a\-m\-e\-t\-e\-r\-s\- & \raggedright A list of the Hagedorn parameter sets $\Pi_i$
          of component $\Phi_i$ of the initial wavepacket
          $\Psi$.

&\\
\cline{1-2}
\raggedright n\-g\-n\- & \raggedright The Number of grid nodes $\gamma_i$ in position
          space.

\textbf{Value:}
{\tt 4096}&\\
\cline{1-2}
\raggedright f\- & \raggedright Scaling factor $f$ for the computational domain
          $\Omega$ in position space. The interval in the
          position space is given by $[-f\pi, f\pi]$.

\textbf{Value:}
{\tt 5.0}&\\
\cline{1-2}
\raggedright b\-a\-s\-i\-s\-\_\-s\-i\-z\-e\- & \raggedright The number $K$ of basis functions $\phi_k$
          used for Hagedorn wavepackets $\Phi$.

\textbf{Value:}
{\tt 64}&\\
\cline{1-2}
\raggedright l\-e\-a\-d\-i\-n\-g\-\_\-c\-o\-m\-p\-o\-n\-e\-n\-t\- & \raggedright The leading component index $\chi$ of the
          eigenvalue $\lambda_\chi$ that
          governs the propagation of the Hagedorn parameters
          $\Pi$ for homogeneous wavepackets.

\textbf{Value:}
{\tt 0}&\\
\cline{1-2}
\raggedright o\-u\-t\-f\-i\-l\-e\-\_\-n\-o\-d\-e\-s\- & \raggedright Filename of the output file that contains the grid nodes
          $\gamma_i$.

\textbf{Value:}
{\tt \texttt{'}\texttt{nodes.dat}\texttt{'}}&\\
\cline{1-2}
\raggedright o\-u\-t\-f\-i\-l\-e\-\_\-w\-a\-v\-e\-f\-u\-n\-c\-t\-i\-o\-n\- & \raggedright Filename of the output file that contains the wavefunction
          $\psi_i$.

\textbf{Value:}
{\tt \texttt{'}\texttt{wavefunction.dat}\texttt{'}}&\\
\cline{1-2}
\raggedright o\-u\-t\-f\-i\-l\-e\-\_\-e\-n\-e\-r\-g\-i\-e\-s\- & \raggedright Filename of the output file that contains the energies $E$.

\textbf{Value:}
{\tt \texttt{'}\texttt{energies.dat}\texttt{'}}&\\
\cline{1-2}
\raggedright o\-u\-t\-f\-i\-l\-e\-\_\-o\-p\-e\-r\-a\-t\-o\-r\-s\- & \raggedright Filename of the output file that contains the operators $T$
          and $V$.

\textbf{Value:}
{\tt \texttt{'}\texttt{operators.dat}\texttt{'}}&\\
\cline{1-2}
\raggedright o\-u\-t\-f\-i\-l\-e\-\_\-p\-a\-r\-a\-m\-e\-t\-e\-r\-s\- & \raggedright Filename of the output file that contains the Hagedorn
          parameters $\Pi_i$.

\textbf{Value:}
{\tt \texttt{'}\texttt{parameters.dat}\texttt{'}}&\\
\cline{1-2}
\raggedright o\-u\-t\-f\-i\-l\-e\-\_\-c\-o\-e\-f\-f\-i\-c\-i\-e\-n\-t\-s\- & \raggedright Filename of the output file that contains the Hagedorn
          coefficients $c^i$.

\textbf{Value:}
{\tt \texttt{'}\texttt{coefficients.dat}\texttt{'}}&\\
\cline{1-2}
\raggedright w\-r\-i\-t\-e\-\_\-n\-t\-h\- & \raggedright Write data to disk only each $n$-th timestep.

\textbf{Value:}
{\tt 1}&\\

\cline{1-2}
\end{longtable}

    \index{Parameters \textit{(module)}|)}
